Model Finding for Recursive Functions in SMT

### 2015-04-29 tex/main.tex
Mention front end.

@@

@@

\section{Model Finding for Recursive Functions in SMT}

@@

\begin{example}
Consider the (combined) signature $\Sigma = \Sigmalia \cup \Sigma_u$,
where $\ssorts{\Sigma_u} = \{ \sortint \}$ and 
$\sfuns{\Sigma_u} = \{ s\typ{ \sortint \rightarrow \sortint }, c\typ{ \sortint } \}$.
Consider the $\Sigma$-formula $\varphi$:
\begin{eqnarray} \label{eq:ex-before}
\forall_{s} x\typ{\sortint}. ite( x \leq 0, s(x) \teq 0, s( x ) \teq x + s( x-1 ) ) \wedge s( c ) > 100
\end{eqnarray} 
Here, $s$ is a function that returns the sum of positive numbers between $0$ and its argument $x$.
The formula $\varphi$ is well-defined with respect to functions $\sfundefs{\Sigma_u}$,
and states that the sum of all positive numbers between $0$ and $c$ is greater than $100$.
This formula is satisfiable with a model that interprets $c$ as an integer greater than or equal to $14$.
Due to the universally quantified formula in the left disjunct,
current SMT techniques~\cite{GeDeM-CAV-09, ReyEtAl-1-RR-13} are unable to find a model for this formula.
The signature $\extendsig{\Sigma}$ includes the type $\fargsort{s}$,
and the uninterpreted function $\farg{s}$ of type $\fargsort{s} \rightarrow \sortint$.
The result of $\conv( \varphi, \ptrue )$ is the pair $( \psi, \emptyset )$, where after simplification, $\psi$ is the $\extendsig{\Sigma}$-formula:
\begin{eqnarray} \label{eq:ex-after}
\begin{split}
\forall y\typ{\fargsort{s}}. ite( & \farg{s}( y ) \leq 0,  \\
 & s(\farg{s}( y )) \teq 0, \\
 & s(\farg{s}( y )) \teq \farg{s}( y ) + s( \farg{s}( y )-1 ) ) \\
 & \wedge \exists z\typ{\fargsort{s}}. \farg{s}( z ) \teq \farg{s}( y )-1 )
\end{split}  
\wedge ( s( c ) > 100 \wedge \exists z\typ{\fargsort{s}}. \farg{s}( z ) \teq c )
\end{eqnarray} 
The universal quantification in the resulting formula is over an uninterpreted type $\fargsort{s}$, 
thus enabling finite model finding techniques~\cite{ReyEtAl-1-RR-13} to search for a finite interpretation for $\fargsort{s}$.
Moreover, all occurrences of $y$ in this formula are beneath applications of the uninterpreted function $\farg{s}$,
implying that the formula is in the essentially uninterpreted fragment of~\cite{GeDeM-CAV-09}, for which there exists a complete instantiation procedure.
We observe that both \cvc and \ziii run indefinitely on formula~(\ref{eq:ex-before}),
and both produce a model for formula~(\ref{eq:ex-after}) in $<0.1$ second.
$\square$
\end{example}

@@

\section{Evaluation}

@@

\begin{figure}
\centering
{
\begin{tabular}{|l|cc|cc|cc|cc|l|cc|cc|cc|cc|}
\hline                                                                
  & \multicolumn{2}{c|}{{\bf z3}}     & \multicolumn{2}{c|}{{\bf cvc4}}     & \multicolumn{2}{c|}{{\bf cvc4+f}}     & \multicolumn{2}{c|}{{\bf cvc4+i}}     & & \multicolumn{2}{c|}{{\bf z3}}     & \multicolumn{2}{c|}{{\bf cvc4}}     & \multicolumn{2}{c|}{{\bf cvc4+f}}     & \multicolumn{2}{c|}{{\bf cvc4+i}}     
\\                                                                    
  & u & s & u & s & u & s & u & s & & u & s & u & s & u & s & u & s 
\\                                                                    
\hline                                                                    
{\bf isa} & 14  & 0 & 15  & 0 & 15  & 0 & {\bf 61}  & 0 & {\bf isa+w} & 15  & 0 & 15  & 0 & 15  & 0 & 14  & 0
\\                                                                  
{\bf leon}  & 73  & 0 & 80  & 0 & 80  & 0 & {\bf 96}  & 0 & {\bf leon+w}  & 78  & 2 & 80  & 0 & 76  & {\bf 9} & 78  & 0
\\                                                                   
{\bf isa+m} & 17  & 0 & 18  & 0 & 18  & 0 & {\bf 44}  & 0 & {\bf isa+mw}  & 18  & 35  & 18  & 0 & 18  & {\bf 153} & 17  & 0
\\                                                                  
{\bf leon+m}  & 83  & 11  & 103 & 6 & 104 & 6 & {\bf 117} & 6 & {\bf leon+mw} & 98  & 75  & 98  & 6 & 95  & {\bf 169} & 98  & 6
\\                                                                  
\hline                                                             
{\bf total} & 187 & 11  & 216 & 6 & 217 & 6 & {\bf 318} & 6 & {\bf total} & 209 & 112 & 211 & 6 & 204 & {\bf 331} & 207 & 6
\\                                                                                            
\hline                                                                                            
\end{tabular}
\\
}
\caption{Number of unsat and sat responses for of all configurations on benchmarks 
before and after the translation $\conv$.}
\label{fig:results}
\end{figure}

@@

We have seen a satisfiability-preserving translation $\conv$ that significantly improves the effectiveness of known SMT techniques
for finding counterexamples for formulas involving recursive function definitions.
For convienience, the translation $\conv$ has been implemented in the latest development version of \cvc~\cite{ReyEtAl-1-RR-13}.

@@

{%\footnotesize
\def\ackname{Acknowledgment}
\paragraph{%\footnotesize
\ackname.}

This research is partially supported by the Inria technological development
action \relax{Contrexemples Utilisables par Isabelle et Coq} (CUIC).

>>>>

\newcommand{\definefunreccmd}{\con{define}\text{-}\con{fun}\text{-}\con{rec}}
\newcommand{\definefunsreccmd}{\con{define}\text{-}\con{funs}\text{-}\con{rec}}

@@

in smt : z3, cvc4 (briefly, text in main section gives detail)

@@

\section{Model Finding for Recursive Functions in SMT}
\label{sec:model-finding-rec-fun}

@@

\begin{example}
\label{ex:translation}
Consider the (combined) signature $\Sigma = \Sigmalia \cup \Sigma_u$,
where $\ssorts{\Sigma_u} = \{ \sortint \}$ and 
$\sfuns{\Sigma_u} = \{ s\typ{ \sortint \rightarrow \sortint }, c\typ{ \sortint } \}$.
Consider the $\Sigma$-formula $\varphi$:
\begin{eqnarray} \label{eq:ex-before}
\forall_{s} x\typ{\sortint}. ite( x \leq 0, s(x) \teq 0, s( x ) \teq x + s( x-1 ) ) \wedge s( c ) > 100
\end{eqnarray} 
Here, $s$ is a function that returns the sum of positive numbers between $0$ and its argument $x$.
The formula $\varphi$ is well-defined with respect to functions $\sfundefs{\Sigma_u}$,
and states that the sum of all positive numbers between $0$ and $c$ is greater than $100$.
This formula is satisfiable with a model that interprets $c$ as an integer greater than or equal to $14$.
Due to the universally quantified formula in the left disjunct,
current SMT techniques~\cite{GeDeM-CAV-09, ReyEtAl-1-RR-13} are unable to find a model for this formula.
The signature $\extendsig{\Sigma}$ includes the type $\fargsort{s}$,
and the uninterpreted function $\farg{s}$ of type $\fargsort{s} \rightarrow \sortint$.
The result of $\conv( \varphi, \ptrue )$ is the pair $( \psi, \emptyset )$, where after simplification, $\psi$ is the $\extendsig{\Sigma}$-formula:
\begin{eqnarray} \label{eq:ex-after}
\begin{split}
\forall y\typ{\fargsort{s}}. ite( & \farg{s}( y ) \leq 0,  \\
 & s(\farg{s}( y )) \teq 0, \\
 & s(\farg{s}( y )) \teq \farg{s}( y ) + s( \farg{s}( y )-1 ) ) \\
 & \wedge \exists z\typ{\fargsort{s}}. \farg{s}( z ) \teq \farg{s}( y )-1 )
\end{split}  
\wedge ( s( c ) > 100 \wedge \exists z\typ{\fargsort{s}}. \farg{s}( z ) \teq c )
\end{eqnarray} 
The universal quantification in the resulting formula is over an uninterpreted type $\fargsort{s}$, 
thus enabling finite model finding techniques~\cite{ReyEtAl-1-RR-13} to search for a finite interpretation for $\fargsort{s}$.
Moreover, all occurrences of $y$ in this formula are beneath applications of the uninterpreted function $\farg{s}$,
implying that the formula is in the essentially uninterpreted fragment of~\cite{GeDeM-CAV-09}, for which there exists a complete instantiation procedure.
We observe that both \cvc and \ziii run indefinitely on formula~(\ref{eq:ex-before}),
and both produce a model for formula~(\ref{eq:ex-after}) in $<0.1$ second.
$\square$
\end{example}

@@

\section{Evaluation}
\label{sec:evaluation}

@@

\begin{figure}[t]
\centering
{
\begin{tabular}{|l|cc|cc|cc|cc|l|cc|cc|cc|cc|}
\hline                                                                
  & \multicolumn{2}{c|}{{\bf z3}}     & \multicolumn{2}{c|}{{\bf cvc4}}     & \multicolumn{2}{c|}{{\bf cvc4+f}}     & \multicolumn{2}{c|}{{\bf cvc4+i}}     & & \multicolumn{2}{c|}{{\bf z3}}     & \multicolumn{2}{c|}{{\bf cvc4}}     & \multicolumn{2}{c|}{{\bf cvc4+f}}     & \multicolumn{2}{c|}{{\bf cvc4+i}}     
\\                                                                    
  & u & s & u & s & u & s & u & s & & u & s & u & s & u & s & u & s 
\\                                                                    
\hline                                                                    
{\bf isa} & 14  & 0 & 15  & 0 & 15  & 0 & {\bf 61}  & 0 & {\bf isa+w} & 15  & 0 & 15  & 0 & 15  & 0 & 14  & 0
\\                                                                  
{\bf leon}  & 73  & 0 & 80  & 0 & 80  & 0 & {\bf 96}  & 0 & {\bf leon+w}  & 78  & 2 & 80  & 0 & 76  & {\bf 9} & 78  & 0
\\                                                                   
{\bf isa+m} & 17  & 0 & 18  & 0 & 18  & 0 & {\bf 44}  & 0 & {\bf isa+mw}  & 18  & 35  & 18  & 0 & 18  & {\bf 153} & 17  & 0
\\                                                                  
{\bf leon+m}  & 83  & 11  & 103 & 6 & 104 & 6 & {\bf 117} & 6 & {\bf leon+mw} & 98  & 75  & 98  & 6 & 95  & {\bf 169} & 98  & 6
\\                                                                  
\hline                                                             
{\bf total} & 187 & 11  & 216 & 6 & 217 & 6 & {\bf 318} & 6 & {\bf total} & 209 & 112 & 211 & 6 & 204 & {\bf 331} & 207 & 6
\\                                                                                            
\hline                                                                                            
\end{tabular}
\\
}
\caption{Number of unsat and sat responses for of all configurations on benchmarks 
before and after the translation $\conv$.}
\label{fig:results}
\end{figure}

@@

We have seen a satisfiability-preserving translation $\conv$ that significantly improves the effectiveness of known SMT techniques
for finding counterexamples for formulas involving recursive function definitions.
For convienience, the translation $\conv$ has been implemented in the latest development version of \cvc~\cite{ReyEtAl-1-RR-13}.
Function definitions $\forallf{f} \vec x. \varphi$ can be written in using the $\definefunreccmd$ command, 
proposed in the draft of SMT version 2.5~\cite{}.
For instance, the input from Example~\ref{ex:translation} could be encoded as:

\begin{verbatim}
(define-fun-rec s ((x Int)) Int (ite (<= x 0) 0 (s (- x 1))))
(assert (> (s x) 100))
(check-sat)
\end{verbatim}

When reading this input, 
\cvc adds the (annotated) quantified formula $\forallf{s} x. s( x ) \teq \lite( x \leq 0, 0, s( x-1 ) )$ to its list of assertions,
which after rewriting becomes $\forallf{s} x. \lite( x \leq 0, s( x ) \teq 0, s( x ) \teq s( x-1 ) )$.
If \cvc is in finite model finding mode for recursive functions is enabled (using the command line parameter ``--fmf-fun"),
then it will replace its list of known assertions based on the conversion $\conv$ before checking for satisfiability.

@@

{%\footnotesize
\def\ackname{Acknowledgment}
\paragraph{%\footnotesize
\ackname.}
We would like to thank Damien Busato-Gaston and Emmanouil Koukoutos for providing the initial set of benchmarks used in the evaluation.
This research is partially supported by the Inria technological development
action \relax{Contrexemples Utilisables par Isabelle et Coq} (CUIC).
